An exposition on the propriety of restricted Boltzmann machines

### 2016-08-18 writing/draft.Rmd
added draft intro

---
title: "An exposition on the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Stephen Vardeman, Daniel Nordman"
date: "August, 2016"
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
---

@@

@@

A restricted Boltzmann machine (RBM) is a specific undirected graphical model specified for discrete random variables, binary being the most common. The architecture of an RBM is specified as having two layers, a hidden ($\mathcal{H}$) and a visible layer ($\mathcal{V}$), with no connections within a layer. An example of this structure is visualized in figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.

@@

#Simulations

>>>>

---
title: "An exposition on the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "August, 2016"
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, specified as having two layers, a hidden and a visible layer with no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning - by treating a hidden layer of one RBM as the visible layer in a second RBM a deep architecture can be created. The method is claimed to have the ability to learn very complex and rich structures in data, making these models attractive for supervised learning. However, the generative behavior of RBMs has largely been unexplored. In this paper, we discuss the relationship between parameter specification and the prevalence of degenerate models as well as an exposition of the difficulties in fitting such highly flexible models in the binary case of RBMs.
---

@@

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBM). (TODO: deep learning, deep RBM references.) A RBM is an undirected graphical model constructed for discrete or continuous random variables, specified as having two layers, a hidden and a visible layer with no conditional dependency within a layer. These models are typically used with success in classification of images. (TODO: classification references.) However, the model properties are largely unexplored in the literature and the fitting methodology remains heuristic based and abstruse. In this paper, we attept to give a thorough understanding of the model and its behavior from the context of statistical model theory as well as explore the possibility of a rigorous fitting methodology. (TODO: statistical model definition) Through our study, we find the RBM model to be deficient in two fundamental ways.

First, the models are unsatisfactory as a conceptualization of how data is generated. Specifically, these models often fail to generate data that resembles realistic data. Additionally, it is evidenced that RBMs easily exhibit instability in the parameter space. In practice, this is seen when a single pixel change in an image results in a wildly different classification (TODO: add ref). We investigate these phenomena in section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, a computationally intractible situation is created quickly when fitting with either maximum likelihood or through Bayesian methods. However, even in models small enough to feasibly fit, there are unique challenges due to the overparameterization of RBMs. In section \ref{bayesian-model-fitting} we try three Bayesian modelling techniques, each with the goal of avoiding parts of the parameter space that yield improper models. As could be expected, with computational complexity comes an increase in accuracy, but at the cost of feasibility. Additionally, because these models are overparametrized, any honest fitting method will reproduce the empirical distribution. Thus, there is no smoothing achieved with a RBM that is fit using a rigourous method. We are therefore very skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.

This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution and shows the connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}. Additionally, there are two accompanying appendices with auxillary proofs relating to the choice of data encoding and parameter behavior during fitting with incomplete training data in section \ref{appendix}.

@@

A restricted Boltzmann machine (RBM) is a specific undirected graphical model specified for discrete or continuous random variables, binary being the most common. In this paper, we consider the binary case of an RBM only. The architecture of an RBM is specified as having two layers, a hidden ($\mathcal{H}$) and a visible layer ($\mathcal{V}$), with no connections within a layer. An example of this structure is visualized in figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.

@@

#Explorations
